# Dataset normalization
Let's normalize a dataset! 

We'll use the `billboard` dataset, as demonstrated in [Hadley Whickham's 2014 paper, *Tidy data*](https://vita.had.co.nz/papers/tidy-data.html) (and made available in the [associated
GitHub repository](https://github.com/hadley/tidy-data/blob/master/data/billboard.csv)).

```{r}
# Load in libraries to make normalization more convenient
library(tidyr)
library(dplyr)

# We'll load the dataframe as `billboard_orig`, and modify it as `billboard`
# as we go.
billboard_orig <- read.csv("billboard.csv", stringsAsFactors = FALSE)
```

This dataset gives the history of the Billboard top hits for the year 2000. It's
initial structure has one row for every song, and **78** additional columns that
track its chart performance, providing a starting date, a peak date, and 76
weeks of chart position info. There is also data on a song's genre and duration.

## Stage 0: Initial form

Some datasets start out somewhat normalized. The "least normalized" kind of data
is "unnormalized form" or **UNF**. A key characteristic of UNF data is *nesting*—
or tables-within-tables—and the "atomic"-ness of data: each row should be a single observation.

This dataset does not have any nested data on the surface. However, the fact that
our dataset uses 72 rows to formulate chart history is, in some ways, a thin 
veneer covering the fact that there's a data atomicity problem: the layout isn't too far
off from us having a column of length-72 vectors, instead of 72 columns for the weekly ranking data.

Instead of 76 rows of data, we can "melt" or "pivot" the data to make the relationships 
atomic (and "mutate" some of the results for clarity):

```{r}

# "Pivot" the table to be "long", having lots of observations
# instead of having lots of columns. (Method via https://tidyr.tidyverse.org/articles/tidy-data.html)
billboard <- billboard_orig %>% 
  pivot_longer(
    x1st.week:x76th.week, 
    names_to = "week", 
    values_to = "rank", 
    values_drop_na = TRUE
  ) %>%
  # We can also "mutate" it to clean up the entries
  mutate(
    week = as.integer(gsub("[a-z]|\\.", "", week)),
    date = as.Date(date.entered) + 7 * (week - 1),
    date.entered = NULL # Remove `date.entered`
  )
```

Now that each row is a single observation, we can move on to further normalization.

## Stage 1: First normal form

First normal form (also abbreviated **1NF**) requires:

1.   No nested data (done now, thanks to our above modification! ✅)
2.   Primary keys exist to uniquely identify rows. Technically, a "primary key" can consist of
multiple columns. (This is called a "composite key" if it happens.) In our case, 
we can see a good reason why this might occur: not every song title is unique! 
One song has multiple performers.
```{r}
# This is using the initial `billboard_orig` df, not the long one
duplicated_songs <- billboard_orig$track[duplicated(billboard_orig$track)]
billboard_orig$artist.inverted[billboard_orig$track %in% duplicated_songs]
```

So the true primary key in this dataset is the artist/track combination. To make
this more concrete, we can create a dataframe that assigns each artist/track combo to
a unique ID number, and use that to reference the Billboard data.

This can be done by creating a new table, and then adding the new IDs with a `join` operation

```{r}
track_ids <- unique(billboard[c("artist.inverted","track")])
track_ids$track_id <- 1:length(track_ids$track)

# Join the data, so IDs end up in the long table
billboard <- left_join(track_ids,billboard, by = c("artist.inverted", "track"))

# Remove now-redundant data
billboard$track <- NULL # Remove track title from `billboard`
billboard$artist.inverted <- NULL # Remove artist from `dataframe`billboard`
```

This isn't strictly necessary to fulfill first normal form, but it'll help us later on. 
1NF? Done NF! ✅

## Stage 2: Second normal form

In addition to the requirements of 1NF, second normal form (**2NF**) requires that, if we
*do* have a composite key, every column needs to depend on the whole key, not just
a part of it. For example, if there were a `nationality` column in our data set, 
that would violate 2NF—nationality only depends on the artist, and is unrelated
to the individual song. In that case, we would need to split things up into an
artist/nationality table.

Since we've already swapped our composite key for a numeric ID key, and
don't have any other problematic dependencies, 2NF is automatically fulfilled! 🎉✅

## Stage 3: Third normal form

Third normal form (**3NF**) looks at column dependencies even further—it requires that
non-primary key columns (also called *candidate keys*) *only* depend on primary keys. For example,
in a table with books as the primary key, the author's birth date doesn't depend
on the book title—it depends on the author, which depends on the title. To resolve 
this, author birth date would need to be moved to an author info table, where the 
authors are the keys. Another way to think about this is a separation of interests or subjects.

We can see in our current table that subjects aren't quite separate: `billboard`
contains information on chart positions, but that's not intrinsically tied to a 
song's `time` or `genre`—so one step we can take is moving those to the `track_id`
dataframe.
```{r}
# Gather up tracks' lengths and genres
id_time_genre <- unique(billboard[c("track_id","time", "genre")])

# Attach them to the `track_ids` dataframe
track_ids <- left_join(track_ids, id_time_genre, by = "track_id")

# Remove the transferred columns
billboard$time <- NULL
billboard$genre <- NULL
```

This is also officially "tidy", under the [tidyverse
definition](https://tidyr.tidyverse.org/articles/tidy-data.html#tidy-data). ✅

## Stage 4: Beyond tidy (fourth normal form+)

While "tidy" data is closely related to noromal forms and fulfills many of the
requirements of the first 3 forms, there is even more normalization that can be done.
There are currently 6 forms that have been defined, but we'll stop at 4 for now. 

Fourth normal form (**4NF**) requires that there are no "non-trivial multivalued
dependencies" in a table—in other words, if a column depends on only one key,
it should live in a table where that key is the primary key, not in a table with
a compound key.

A telltale sign of a 4NF violation is redundant or duplicate data. Looking at our
dataframe, there is one column where this is still the case: `date.peaked`. This is
because, in the `billboard` table, each row is identified uniquely by a `track_id`
and a `week` (or `date`), but `date.peaked` actually really only depends on the
track—so it should also be moved to the other table. 

```{r}
# Gather up `track_id`s and their `date_peaked` info
id_peaks <- unique(billboard[c("track_id","date.peaked")])

# Attach them to the `track_ids` dataframe
track_ids <- left_join(track_ids, id_peaks, by = "track_id")

# Remove the transferred columns
billboard$date.peaked <- NULL
```

Now our redundancy has been removed, and we're in 4NF—*beyond* "tidy"! ✅